\documentclass{article}

\usepackage{dsfont}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{semantic}
\usepackage{xcolor}
\usepackage{listings}
\lstset{
basicstyle=\ttfamily,
frame=single,
numbers=left
}
\usepackage{fullpage}
\usepackage{mathpartir}

\newcommand{\todo}[1]{\textcolor{red}{\textbf{ToDo}\ifx&#1&\else ~[\emph{#1} ]~\fi}}
\newcommand{\fixme}[1]{\textcolor{purple}{\textbf{FixMe}\ifx&#1&\else ~[ \emph{#1} ]~\fi}}
\newcommand{\note}[1]{\textcolor{blue}{\textbf{Note}\ifx&#1&\else ~[ \emph{#1} ]~\fi}}

\title{SOS --- Homework}

\author{Simon Bihel \and Lesly-Ann Daniel}

\date{M2 SIF, 2017--2018}

\begin{document}
\maketitle

\section*{Question 1 - v1}
%------------------------ Part 1
Let \(\mathcal{N}\) be the interpretation function for constants:
\[ \mathcal{N} = Num \rightarrow \mathbb{Z} \]
In our extended \texttt{While} language, values are either integers or arrays, thus we define the set of program values:
\[ \textbf{Val} = \mathbb{Z} \cup \textbf{Arr}\]
We define the set of arrays as the set of all functions from indexes (natural numbers) to program values, plus the value \(nil\) for undefined arrays: \note{We could define the set of array as the set of finite sequences of values, it might be more readable/easier to understand...}
\[\textbf{Arr} = \{x:\mathbb{N} \rightarrow \textbf{Val}\} \cup \{nil\} \]
Finally, a state is represented by a function from program variables to values:
\[\sigma \in \textbf{State} = \textbf{Var} \rightarrow \textbf{Val} \]

%------------------------ Part 2
Arithmetic expressions are extended from integers to integers or arrays.
The denotational semantic of arithmetic expressions is defined by the following rules:
\begin{center}
\begin{tabular}{ l c l }
\multicolumn{3}{l}{\(\mathcal{A}: \textbf{Aexp} \rightarrow \textbf{State} \rightarrow \textbf{Val}\)} \\
 \(\mathcal{A}\llbracket n \rrbracket \sigma \) & \(=\) & \(\mathcal{N}\llbracket n \rrbracket\)\\ 
 \(\mathcal{A}\llbracket x \rrbracket \sigma \) & \(=\) & \(\sigma x\)\\  
 \(\mathcal{A}\llbracket a_1 \circ a_2 \rrbracket \sigma \) & \(=\) &
    if \(\mathcal{A}\llbracket a_1\rrbracket \sigma \in \mathbb{Z}\) and \(\mathcal{A}\llbracket a_2\rrbracket \sigma \in \mathbb{Z}\)\\
    & & then \(\mathcal{A}\llbracket a_1\rrbracket \sigma \circ \mathcal{A}\llbracket a_2\rrbracket \sigma\)\\ 
 \(\mathcal{A}\llbracket x.length \rrbracket \sigma \) & \(=\) & if \(\sigma x \in \{\textbf{Arr} \setminus {nil}\}\) then \(|dom(\sigma x)|\) \\ 
 \(\mathcal{A}\llbracket x[a] \rrbracket \sigma \) & \(=\) &
    if \(\sigma x \in \{\textbf{Arr} \setminus {nil}\}\) and \(\mathcal{A}\llbracket a\rrbracket \sigma \in dom(\sigma x)\)\\
    & & then \(\sigma x (\mathcal{A}\llbracket a\rrbracket \sigma)\)
\end{tabular}
\end{center}
Note the following changes from the definition of the while language:
\begin{itemize}
\item \(\mathcal{A}\llbracket x \rrbracket \sigma \) can now return an integer or an array.
\item \(\mathcal{A}\llbracket a_1 \circ a_2 \rrbracket \sigma \) is not defined if \(a_1\) and \(a_2\) are not integers.
\item \(\mathcal{A}\llbracket x.length \rrbracket \sigma \) is not defined if \(x\) is an integer or the \(null\) array.
\item \(\mathcal{A}\llbracket x.length \rrbracket \sigma \) is the cardinal of the domain of \(\sigma x\) otherwise, i.e. the number of indexes upon which the array \(x\) is defined.
\item \(\mathcal{A}\llbracket x[a] \rrbracket \sigma \) is not defined if \(x\) is an integer or the \(null\) array. Moreover, it is only defined if the index \(a\) is in the domain of \(\sigma x\) i.e. if \(a\) is a valid index of \(x\).
\item If \(\mathcal{A}\llbracket x[a] \rrbracket \sigma \) is defined, it first applies \(x\) to \(\sigma\) which returns the array, represented as a function. Then it does the dereferencing by applying \(\mathcal{A}\llbracket a \rrbracket \sigma\) to the array function, and returns an integer or an array.
\end{itemize}

The denotational semantic of boolean expressions is defined by the following rules:
\begin{center}
\begin{tabular}{ l c l }
\multicolumn{3}{l}{\(\mathcal{B}: \textbf{Bexp} \rightarrow \textbf{State} \rightarrow \mathbb{B}\)} \\
 \(\mathcal{B}\llbracket true \rrbracket \sigma \) & \(=\) & \(\textbf{tt}\)\\ 
 \(\mathcal{B}\llbracket false \rrbracket \sigma \) & \(=\) & \(\textbf{ff}\)\\  
 \(\mathcal{B}\llbracket a_1 \circ a_2 \rrbracket \sigma \) & \(=\) &
    if \(\mathcal{A}\llbracket a_1\rrbracket \sigma \in \mathbb{Z}\) and \(\mathcal{A}\llbracket a_2\rrbracket \sigma \in \mathbb{Z}\)\\
    & & then \(\mathcal{A}\llbracket a_1\rrbracket \sigma \circ \mathcal{A}\llbracket a_2\rrbracket \sigma\)\\ 
 \(\mathcal{B}\llbracket \text{not } b \rrbracket \sigma \) & \(=\) & \(\neg(\mathcal{B}\llbracket b \rrbracket \sigma)\) \\
 \(\mathcal{B}\llbracket b_1 \text{ and } b_2 \rrbracket \sigma \) & \(=\) &
    \((\mathcal{B}\llbracket b_1 \rrbracket \sigma) \wedge (\mathcal{B}\llbracket b_2 \rrbracket \sigma)\) \\
 \(\mathcal{B}\llbracket b_1 \text{ or } b_2 \rrbracket \sigma \) & \(=\) &
    \((\mathcal{B}\llbracket b_1 \rrbracket \sigma) \vee (\mathcal{B}\llbracket b_2 \rrbracket \sigma)\) \\
\end{tabular}
\end{center}
Note that  \(\mathcal{B}\llbracket a_1 \circ a_2 \rrbracket \sigma \) is not defined if \(a_1\) and \(a_2\) are not integers.
Consequently, some rules parameterized by booleans expressions in the small-step semantic might be blocking on undefined booleans expressions.

The structural operational semantics (small-step) comprises all the previous rules plus the following extension:
\begin{mathpar}
\inference[ARR\_DECL\_I]{}{(x := \text{new array[int]}(a), \sigma) \rightarrow
\sigma[x \mapsto x(i) = 0, \forall i \in [0 .. (a-1)] ]}
\and
\inference[ARR\_DECL\_A]{}{(x := \text{new array[} \text{\textit{arr\_dcl}]}(a), \sigma) \rightarrow
\sigma[x \mapsto x(i) = nil, \forall i \in [0 .. (a-1)] ]}
\and
\inference[ARR\_UPDATE]{}
    {(x[a_1] := a_2, \sigma) \rightarrow
     \sigma[x \mapsto \sigma x[\mathcal{A}\llbracket a_1 \rrbracket \sigma \mapsto \mathcal{A}\llbracket a_2 \rrbracket \sigma]]}
     [if \(\mathcal{A}\llbracket a_1\rrbracket \sigma \in dom(\sigma x)\)]
\end{mathpar}
The array declaration is split into two rules:
\begin{itemize}
 \item \textsc{Arr\_Decl\_I} is the declaration of arrays of integers. The arrays \(x\) is initialized as a function mapping indexes (from 0 to \(a-1\)), to 0.
 \item \textsc{Arr\_Decl\_A} is the declaration of arrays of arrays. The arrays \(x\) is initialized as a function mapping indexes (from 0 to \(a-1\)), to \(nil\).
\end{itemize}
The \textsc{Arr\_Update} is conditioned by the fact that the index \(a_1\) is in the bound of the array. \textsc{Arr\_Update} consists in updating the function \(\sigma x\) by mapping the index \(a_1\) to its new value \(a_2\), and then updating the store \(\sigma\) with the new value of \(\sigma x\).


\section*{Question 2}
The definition of the big-step semantic would not change, except for the addition of the above rules that would be straightforward.
Indeed, the rules defined above do not manipulate commands, so their translation into big-step semantic only consists in changing the transition relation from small-step to big-step.

There would be more blocking configurations because some rules of the big step-semantic are parameterized by a boolean evaluation which can be undefined (to avoid the comparison of arrays) in our extended semantic contrary to the original semantic.

\section*{Question 3}
% Question 3. We suppose that all instructions are given a unique label. Propose a data-flow analysis for
% determining the potential may-aliases at each program point. You may want to use the allocation site of
% an array as abstraction but other abstract domains for modeling memory allocation are also welcome. The
% allocation site of an array is the label of the new instruction that created it.
In this section, we propose a dataflow analysis for determining the potential may-aliases at each program point. We use the allocation site of array as an abstraction, which we represent as the label of the instruction that allocated it.

\paragraph{Control Flow}
The syntax of the \texttt{While} language is extended with labels, representing program points:
\begin{equation*} \label{eq1}
\begin{split}
S::= & [x := a]^l
     \ |\ [\text{skip}]^l
     \ |\ S_1;S_2
     \ |\ \text{if } [b]^l \text{ then } S_1 \text{ else } S_2 
     \ |\ \text{while } [b]^l \text{ do } S \ |\\
     & [x := \text{new } arr\_decl(a)]^l
     \ |\ [x[a_1] := a_2]^l\\
& n \in \textbf{Num}, x \in \textbf{Var}, a \in \textbf{Aexp}, b \in \textbf{Bexp}, S \in \textbf{Comm}, l \in \textbf{Lab}
\end{split}
\end{equation*}
For each instruction \(S \in \textbf{Comm}\), we associate the following functions to represent the control flow graph of the program:
\begin{center}
\begin{tabular}{ l l l }
\(init(S)\) & \(\in \textbf{Lab}\) &  \textit{the entry point label of S}\\
\(final(S)\) & \(\subseteq \textbf{Lab}\) &  \textit{the exit point labels of S}\\
\(labels(S)\) & \(\subseteq \textbf{Lab}\) &  \textit{the labels which appear in S}\\
\(flow(S)\) & \(\subseteq \textbf{Lab} \times \textbf{Lab}\) &  \textit{the edges of the control flow graph}
\end{tabular}
\end{center}
which are defined as follow\footnote{Note that these definitions are just an extension of the definitions presented in the course with the new commands to manipulate arrays.}:
\begin{center}
\begin{tabular}{ l l }
\( init([x := a]^l) \)                                           & \( = l\)\\
\( init([\text{skip}]^l) \)                                      & \( = l\)\\
\( init(S_1;S_2) \)                                              & \( = init(S_1)\)\\
\( init(\text{if } [b]^l \text{ then } S_1 \text{ else } S_2) \) & \( = l\)\\
\( init(\text{while } [b]^l \text{ do } S) \)                    & \( = l\)\\
\( init([x := \text{new } arr\_decl(a)]^l) \)                    & \( = l\)\\
\( init([x[a_1] := a_2]^l) \)                                    & \( = l\)
\end{tabular}
\end{center}
\begin{center}
\begin{tabular}{ l l }
\( final([x := a]^l) \)                                           & \( = \{l\}\)\\
\( final([\text{skip}]^l) \)                                      & \( = \{l\}\)\\
\( final(S_1;S_2) \)                                              & \( = final(S_2)\)\\
\( final(\text{if } [b]^l \text{ then } S_1 \text{ else } S_2) \) & \( = final(S_1) \cup final(S_2)\)\\
\( final(\text{while } [b]^l \text{ do } S) \)                    & \( = \{l\}\)\\
\( final([x := \text{new } arr\_decl(a)]^l) \)                    & \( = \{l\}\)\\
\( final([x[a_1] := a_2]^l) \)                                    & \( = \{l\}\)
\end{tabular}
\end{center}
\begin{center}
\begin{tabular}{ l l }
\( labels([x := a]^l) \)                                           & \( = \{l\}\)\\
\( labels([\text{skip}]^l) \)                                      & \( = \{l\}\)\\
\( labels(S_1;S_2) \)                                              & \( = labels(S_1) \cup labels(S_2)\)\\
\( labels(\text{if } [b]^l \text{ then } S_1 \text{ else } S_2) \) & \( = \{l\} \cup labels(S_1) \cup labels(S_2)\)\\
\( labels(\text{while } [b]^l \text{ do } S) \)                    & \( = \{l\} \cup labels(S)\)\\
\( labels([x := \text{new } arr\_decl(a)]^l) \)                    & \( = \{l\}\)\\
\( labels([x[a_1] := a_2]^l) \)                                    & \( = \{l\}\)
\end{tabular}
\end{center}
\begin{center}
\begin{tabular}{ l l }
\( flow([x := a]^l) \)                                           & \( = \emptyset \)\\
\( flow([\text{skip}]^l) \)                                      & \( = \emptyset \)\\
\( flow(S_1;S_2) \)                                              &
    \( = flow(S_1) \cup flow(S_2) \cup \{(l,init(S_2))\ |\ l \in final(S_1)\}\)\\
\( flow(\text{if } [b]^l \text{ then } S_1 \text{ else } S_2) \) &
    \( = flow(S_1) \cup flow(S_2) \cup \{(l,init(S_1))\} \cup \{(l,init(S_2))\}\)\\
\( flow(\text{while } [b]^l \text{ do } S) \)                    &
    \( = flow(S) \cup \{(l',l)\ |\ l' \in final(S)\} \cup \{(l,init(S))\}\)\\
\( flow([x := \text{new } arr\_decl(a)]^l) \)                    & \( = \emptyset \)\\
\( flow([x[a_1] := a_2]^l) \)                                    & \( = \emptyset \)\\
\end{tabular}
\end{center}

\paragraph{May-Aliases Analysis}
In our may-aliases (MA) analysis, the allocation site of an array is used as an abstraction; for each array we keep track of their ``reachable'' allocation sites.
An allocation is represented by a couple \((v,l) \in \textbf{Var} \times \textbf{Lab}^\textbf{?}\)
with \(\textbf{Lab}^\textbf{?} = \textbf{Lab} \cup \{?\}\), such that:
\begin{itemize}
 \item \((v,l)\) : The array \(v\) has been allocated a program point \(l\).
 \item (v,?) : The array \(v\) is not allocated.
\end{itemize}
At each label program point \(l\), we compute the sets \(M\!A_\text{in}(l), M\!A_\text{out}(l) \in \mathcal{P}(\textbf{Var} \times \textbf{Lab}^?)\) such that:
\begin{itemize}
 \item \(M\!A_\text{in}(l)\), the allocated arrays that enter in \(l\)
 \item \(M\!A_\text{out}(l)\), the allocated arrays that exit from \(l\)
\end{itemize}
Finally, we define the relation \(x \sim_{M\!A(l)} x'\) stating that two memory access via \(x\) and \(x'\) may have the same target at point \(l\), such that:
\[x \sim_{M\!A(l)} x' \iff [x = x' \quad \vee \quad (\exists l'.\ (x,l') \in M\!A(l) \wedge (x',l') \in M\!A(l)) ]
\]

The ``gen-kill'' rules for the MA analysis define which allocations are created and which are killed at each command:
\begin{center}
\begin{tabular}{ l l l l}
\(1.\ kill([x := a]^l) \)      & = & \( \{(x,l')\ |\ l' \in \textbf{Lab}^?\} \) & \textit{kill any previous alloc. of x}\\ 
\(2.\ kill([\text{skip}]^l) \) & = & \( \emptyset \) \\
\(3.\ kill([b]^l) \)           & = & \( \emptyset \) \\
\(4.\ kill([x := \text{new } arr\_decl(a)]^l) \) & = & \( \{(x,l')\ |\ l' \in \textbf{Lab}^?\} \) & \textit{kill any previous alloc. of x}\\ 
\(5.\ kill([x[a_1] := a_2]^l) \) & = & \(\emptyset\)
\end{tabular}
\end{center}
In the rule 1, we assign a new value to \(x\), therefore any previous allocation of \(x\) becomes non-reachable and all its allocation sites are removed from MA; the same reasoning holds for rule 4.
In rule 5, we would want to remove from MA the previous allocation sites reachable from \(x[a_1]\), however in our analysis they cannot be distinguished from the allocation sites reachable from \(x[a_2]\) or even from \(x\). Therefore we do nothing and keep these potentially non-reachable allocation sites in MA, and get an over approximation.
\begin{center}
\begin{tabular}{ l l l }
\(1.\ gen([x_1 := x_2]^l) \)    & =  \( \{(x_1,l')\ |\ (x_2,l') \in  M\!A_\text{in}(l)\} \)
                             & \textit{add any allocation reachable via \(x_2\)}\\
\(2.\ gen([x_1 := x_2[a]]^l) \) & =  \( \{(x_1,l')\ |\ (x_2,l') \in  M\!A_\text{in}(l) \}\)
                             & \textit{add any allocation reachable via \(x_2\) }\\ 
\(3.\ gen([x := a]^l) \)        & =  \( \emptyset \)
                             & \textit{if \(a_2\) is not a variable}\\ 
\(4.\ gen([\text{skip}]^l) \)   & =  \( \emptyset \) \\
\(5.\ gen([b]^l) \)             & =  \( \emptyset \) \\

\(6.\ gen([x := \text{new } arr\_decl(a)]^l) \) & =  \( \{(x,l)\} \) & \textit{create new alloc. for x}\\

\(7.\ gen([x_1[a] := x_2]^l) \)        & =  \( \{(x',l')\ |\ x_1 \sim_{M\!A_{\text{in}}(l)} x' \wedge (x_2,l') \in  M\!A_\text{in}(l) \}\)
                                    & \textit{\(x_2\) is reachable via any alias of \(x_1\)}\\
\(8.\ gen([x_1[a_1] := x_2[a_2]]^l) \) & =  \( \{(x',l')\ |\ x_1 \sim_{M\!A_{\text{in}}(l)} x' \wedge (x_2,l') \in  M\!A_\text{in}(l) \}\)
                                    & \textit{\(x_2\) is reachable via any alias of \(x_1\)}\\ 
\(9.\ gen([x[a_1] := a_2]^l) \)        & =  \( \emptyset \)
                                    & \textit{if \(a_2\) is not a variable}\\
\end{tabular}
\end{center}
In the rule 1, any allocation site previously reachable via \(x_2\) becomes also reachable via \(x_1\); the same reasoning holds for rule 2. 
In the rule 6, we allocate a fresh array to the variables \(x\), thus we add the current label in MA as the allocation site of \(x\).
In the rules 7 and 8, any allocation site previously reachable via \(x_2\) becomes also reachable via \(x_1\) but also via all the aliases of \(x_1\). Here, we make an over-approximation because we do not differentiate the allocation sites reachable from \(x_1\) from those reachable from \(x_1[a]\) or those reachable from \(x_1[a']\).

Finally, we compute the sets \(M\!A_\text{in}(l)\) and \(M\!A_\text{out}(l)\) for all program point \([i]^l \in P\) :
 \[
    M\!A_\text{in}(l) =
    \begin{cases}
      \{(x,?)\ |\ x \in \textbf{Var}\},                        & \text{if } l=init(P)\\
      \bigcup\limits_{(l,l') \in flow(P)} M\!A_\text{out}(l'), & \text{otherwise}
    \end{cases}
\]
At the beginning of the program, all the variables are set to unallocated. Since MA analysis is a forward analysis
\[
    M\!A_\text{out}(l) = M\!A_\text{in}(l) \setminus kill([i]^l) \cup gen([i]^l)
\]
We can note that MA is a {\em forward} analysis, defining something that {\em may} happen: we are considering the least-fixpoint.
The least fixpoint exists because the functions \(M\!A_\text{in}(l)\) and \(M\!A_\text{out}(l)\) are monotonically increasing.
Moreover, the fixpoint iteration terminates because the lattice used for MA (\(\mathcal{P}(\textbf{Var} \times \textbf{Lab}^?), \subseteq, \cup\)) is finite.

\section*{Question 4}
The MA analysis defined in Question 3 is subject to some imprecisions. It is an over-approximation of the real behavior of the program and thus contains false positive.

First, some infeasible paths might cause a may alias which is impossible in practice.
For instance, the MA analysis of the program in Listing \ref{code:infeasible} detects a may-alias between \(z\) and \(y\) when both conditional statements are evaluated to true, which is not possible in practice.
\todo{Find an example with a program where the type of variables cannot change ?}
\begin{lstlisting}[caption=Example of infeasible execution path causing a may aliasis., label=code:infeasible]
z = 0;
if(x < 0)
  z = new array[int](10);
if(x > 0)
  y = z; // Triggers a may alias but actually there is none.
\end{lstlisting}

Second, our analysis do not differentiate allocations reachable via \(x\) and allocations reachable via \(x[i]\); in both cases, we consider that the allocation site is reachable via \(x\) and add \((x,l)\).
Moreover in case of an assignment \([x_1 := x_2[a]]^l\), we add to \(x_1\) all the reachable assignments from to \(x_2\), although we would like to add only those reachable from \(x_2[a]]^l\).
This constitutes an over-approximation in the analysis.

Finally, when updating an array with a reference \([x[a_1] := a_2]^l\) might override the allocation reachable from \(x[a_1]\) which would be no longer reachable via \(x\), but we cannot identify them and remove them from MA.

\todo{Improvements:}


\section*{Question 5}
\paragraph{Definition of types}
The definition of types we consider in our program are booleans, integers and multidimensional arrays of integers. We define a particular type for the commands. Therefore, we define the types \(T\) as:
\begin{equation*} \label{eq1}
\begin{split}
T' & ::= \text{Nat}\ |\ \text{Array}[T']\\
T  & ::= \text{Bool}\ |\ \text{Cmd}\ |\ T'
\end{split}
\end{equation*}

\paragraph{Guarantees}
The typing rules are constructive and guarantee that the type of variables cannot change.
If there exists a \(\Gamma\) such that the program is well-typed, the following properties are guaranteed on its execution:
\begin{itemize}
 \item The rule \textsc{T-AOp} guarantees that an arithmetic operation can only be applied on integers and not on arrays.
 \item The rule \textsc{T-Len} guarantees that the \texttt{length} operator can only be applied to arrays. It does not guarantee that the reference is not \(nil\), for instance in the program in Listing \ref{code:nilderef}.
 \item The rule \textsc{T-Deref} guarantees that only arrays can be dereferenced and only with an integer. It does not guarantee that the reference is not \(nil\) and that the access lies within the bounds of the array.
 \item The rule \textsc{T-Comp} guarantees that only integers can be compared.
 \item The rule \textsc{T-Assign} guarantees that the type of the variables cannot change, and that only expressions with the same type as the variable can be assigned. 
 \item The rule \textsc{T-ADecl} guarantees that the size of the allocated array is an integer.
 \item The rule \textsc{T-AUpdate} guarantees that the variables dereferenced is an array and that both side of the assignment have the same type. It does not guarantee that the reference is not \(nil\) and that the access lies within the bounds of the array.
\end{itemize}

\begin{lstlisting}[caption=Example of program dereferencing a nil reference., label=code:nilderef]
x := new array[array[int]](10);  // x : Array[Array[Nat]]
y := x[3];                       // y : Array[Nat]
z := y.length                    // Error because y = nil

\end{lstlisting}

We cannot guarantee that a well-typed program is non-blocking.
The first reason is because we initialize arrays of arrays to \(nil\) and it is not possible to guarantee that a reference is not \(nil\) with a type system, it would require a dataflow analysis.
Therefore we cannot guarantee that the program does not try to dereference a variables set to \(nil\). This check is only performed at runtime in the evaluation of arithmetic expressions and then the execution is  blocked.

The second reason is that we cannot guarantee the absence of  array overflows or underflows, providing this guarantee would also require a dataflow analysis.
Therefore, even if we choose to not consider \(nil\) references and initialize arrays of arrays directly (\texttt{new array[array[int]](10)} would then return an \(10 \times 10\) array), we could not have the guarantee that the program is non blocking.

\paragraph{Typing Rules}
We define the type judgment \(\Gamma \vdash S\) meaning that the command \(S\) is well-typed in the context \(\Gamma\) with the following typing rules:

First we define the typing rules for values:
\begin{mathpar}
\textsc{T-Nat} \quad \Gamma \vdash n : \text{Nat}
\and
\textsc{T-True} \quad \Gamma \vdash true : \text{Bool}
\and
\textsc{T-False} \quad \Gamma \vdash false : \text{Bool} 
\end{mathpar}

Then, the typing rules for arithmetic expressions:
\begin{mathpar}
\inference[\textsc{T-Var}]
    {x : t \in \Gamma}
    {\Gamma \vdash x : t}{}
\and
\inference[\textsc{T-AOp}]
    {\Gamma \vdash a_1 : \text{Nat} \quad \Gamma \vdash a_2 : \text{Nat}}
    {\Gamma \vdash a_1 \circ a_2 : \text{Nat}}{(\circ \in \{+, -,\times\})}
\and
\inference[\textsc{T-Len}]
    {x : \text{Array}[t] \in \Gamma}
    {\Gamma \vdash x.length : \text{Nat}}{}
\and
\inference[\textsc{T-Deref}]
    {x : \text{Array}[t] \in \Gamma \quad \Gamma \vdash a : \text{Nat}}
    {\Gamma \vdash  x[a] : t}{}
\end{mathpar}

Then the typing rules for boolean expressions:

\begin{mathpar}
\inference[\textsc{T-Comp}]
    {\Gamma \vdash a_1 : \text{Nat} \quad \Gamma \vdash a_2 : \text{Nat}}
    {\Gamma \vdash a_1 \circ a_2 : \text{Bool}}{(\circ \in \{=, \leq\})}
\and
\inference[\textsc{T-Not}]
    {\Gamma \vdash b : \text{Bool}}
    {\Gamma \vdash \texttt{not } b : \text{Bool}}{}
\and
\inference[\textsc{T-Or}]
    {\Gamma \vdash b_1 : \text{Bool} \quad \Gamma \vdash b_2 : \text{Bool}}
    {\Gamma \vdash  b_1 \texttt{ or } b_2 : \text{Bool}}{}
\and
\inference[\textsc{T-And}]
    {\Gamma \vdash b_1 : \text{Bool} \quad \Gamma \vdash b_2 : \text{Bool}}
    {\Gamma \vdash  b_1 \texttt{ and } b_2 : \text{Bool}}{}
\end{mathpar}

Finally, we define the typing rules for commands:
\begin{mathpar}
\inference[\textsc{T-Assign}]
    {x : t \in \Gamma \quad \Gamma \vdash a : t}
    {\Gamma \vdash  x := a : \text{Cmd}}{}
\and
\inference[\textsc{T-Skip}]
    {}
    {\Gamma \vdash \texttt{skip} : \text{Cmd}}{}
\and
\inference[\textsc{T-Seq}]
    {\Gamma \vdash S_1 : \text{Cmd} \quad \Gamma \vdash S_2 : \text{Cmd}}
    {\Gamma \vdash S_1; S_2 : \text{Cmd}}{}
\and
\inference[\textsc{T-If}]
    {\Gamma \vdash b : Bool \quad \Gamma \vdash S_1 : \text{Cmd} \quad \Gamma \vdash S_2 : \text{Cmd}}
    {\Gamma \vdash \texttt{if } b \texttt{ then } S_1 \texttt{ else } S_2 : \text{Cmd}}{}
\and
\inference[\textsc{T-While}]
    {\Gamma \vdash b : Bool \quad \Gamma \vdash S : \text{Cmd}}
    {\Gamma \vdash \texttt{while } b \texttt{ do } S : \text{Cmd}}{}
\and
\inference[\textsc{T-ADecl-Int}]
    {}
    {\Gamma \vdash\texttt{array[int]} : \text{\text{Array}}[\text{Nat}]}{}
\and
\inference[\textsc{T-ADecl-Arr}]
    {\Gamma \vdash array\_decl : t}
    {\Gamma \vdash \texttt{array}[array\_decl] : \text{\text{Array}}[t]}{}
\and
\inference[\textsc{T-ADecl}]
    {x : t \in \Gamma \quad \Gamma \vdash arr\_decl : t \quad \Gamma \vdash a : \text{Nat}}
    {\Gamma \vdash x := \texttt{new }arr\_decl(a) : \text{Cmd}}{}
\and
\inference[\textsc{T-AUpdate}]
    {x : \text{Array}[t] \in \Gamma \quad \Gamma \vdash a_1 : \text{Nat} \quad \Gamma \vdash a_2 : t}
    {\Gamma \vdash x[a_1] := a_2 : \text{Cmd}}{}
\end{mathpar}
\end{document}
